---
title: "Programming fundamentals"
output: html_notebook
---

```{r Library}
library(dplyr)
library(ggplot2)
library(tidyverse)
```

#1.6 Data Types

##Exercise: Checking Structures
```{r}
# Assign variable values
site_name <- "Heron_Island"
transect_depth_m <- 12.5
bleaching_present <- TRUE
```

```{r}
# Check using function class()
class(site_name)
class(transect_depth_m)
class(bleaching_present)
```
```{r}
# Check using function str()
str(site_name)
str(transect_depth_m)
str(bleaching_present)
```
##Exercise: rounding numbers
```{r}
# Tracking the age of an old-growth Porites coral colony
years_old <- 25.765
```

```{r}
# Clean this up for our summary report
round(years_old, 2)
```

##Manipulating outputs

```{r}
# Make variables
years_old <- 25.765
rounded_age <- round(years_old, 2)

# Combine text and data variables
paste("Average colony age is", rounded_age, "years old")
```
Task: Modify the arguments inside the round() function above so that the mathematical truncation matches the expected field report text exactly.

```{r}
rounded_age <- round(years_old, 1)

# Combine text and data variables
paste("Average colony age is", rounded_age, "years old")
```

#1.7 Data structures


```{r}
fish_lengths <- c(124, 152, 98, 221, 146)
coral_spp <- c("Porites", "Acropora", "Montastrea")
```

```{r}
notes <- c("Acropora", 27.5, TRUE)
```

```{r}
notes <- list("Acropora", 27.5, TRUE)
notes
```
The great thing about lists and vectors is that you can retrieve individual elements using double square brackets to reference their index.
```{r}
notes[[1]]
```

## Data frames and tibbles

```{r creating df}
my_dataframe <- data.frame(no=c(1,2,3),c("Plectropomus","Scarus","Pomacentrus"),c(TRUE,FALSE,TRUE))
my_dataframe
str (my_dataframe)
```

```{r}
my_dataframe$no = as.factor(my_dataframe$no)
str (my_dataframe)
```
A factor is a categorical type, so here we are telling R that column 1 is actually storing a category (a group membership, say a numeric code for which team each of these people are in) rather than a real continuous number. 

#1.8 R Projects and workspace architecture

```{r}

```

# MB5370_Module_1_Coding/
│
├── MB5370_Module_1_Coding.Rproj
├── data/       <- For raw, unedited data files (CSVs, shapes, etc.)
├── code/    <- For dirty/heavy data manipulation code
├── output/    <- For processed tables, figures, and saved objects
└── docs/       <- For your rendered write-ups and final documents
