---
title: "Programming fundamentals"
output:
  html_document:
    df_print: paged
---

```{r Library}
library(dplyr)
library(ggplot2)
library(tidyverse)
```

#Workshop 1

#1.6 Data Types

##Exercise: Checking Structures
```{r}
# Assign variable values
site_name <- "Heron_Island"
transect_depth_m <- 12.5
bleaching_present <- TRUE
```

```{r}
# Check using function class()
class(site_name)
class(transect_depth_m)
class(bleaching_present)
```
```{r}
# Check using function str()
str(site_name)
str(transect_depth_m)
str(bleaching_present)
```
##Exercise: rounding numbers
```{r}
# Tracking the age of an old-growth Porites coral colony
years_old <- 25.765
```

```{r}
# Clean this up for our summary report
round(years_old, 2)
```

##Manipulating outputs

```{r}
# Make variables
years_old <- 25.765
rounded_age <- round(years_old, 2)

# Combine text and data variables
paste("Average colony age is", rounded_age, "years old")
```
Task: Modify the arguments inside the round() function above so that the mathematical truncation matches the expected field report text exactly.

```{r}
rounded_age <- round(years_old, 1)

# Combine text and data variables
paste("Average colony age is", rounded_age, "years old")
```

#1.7 Data structures
```{r}
fish_lengths <- c(124, 152, 98, 221, 146)
coral_spp <- c("Porites", "Acropora", "Montastrea")
```

```{r}
notes <- c("Acropora", 27.5, TRUE)
```

```{r}
notes <- list("Acropora", 27.5, TRUE)
notes
```
The great thing about lists and vectors is that you can retrieve individual elements using double square brackets to reference their index.
```{r}
notes[[1]]
```

## Data frames and tibbles

```{r creating df}
my_dataframe <- data.frame(no=c(1,2,3),c("Plectropomus","Scarus","Pomacentrus"),c(TRUE,FALSE,TRUE))
my_dataframe
str (my_dataframe)
```

```{r}
my_dataframe$no = as.factor(my_dataframe$no)
str (my_dataframe)
```
A factor is a categorical type, so here we are telling R that column 1 is actually storing a category (a group membership, say a numeric code for which team each of these people are in) rather than a real continuous number. 

#1.8 R Projects and workspace architecture

# MB5370_Module_1_Coding/
│
├── MB5370_Module_1_Coding.Rproj
├── data/       <- For raw, unedited data files (CSVs, shapes, etc.)
├── code/    <- For dirty/heavy data manipulation code
├── output/    <- For processed tables, figures, and saved objects
└── docs/       <- For your rendered write-ups and final documents

dir.create("data")
dir.create("code")
dir.create("output")
dir.create("docs")

note to self: added the above into the console to create subfolders.

#Workshop 2


#2.16	Load packages and data

##using the built-in dataset 
```{r}
mpg
```

#2.17 Create your first ggplot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Engine size = displ
Fuiel efficiency = hwy

the plot shows a negative relationship between engine size and fuil efficiency

#2.18 Understand the ‘grammar of graphics’
##2.18.1 Graphing template

##2.18.2 Aesthetic mappings
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))#using size,sizes up the geom_points
#> Warning: Using size for a discrete variable is not advised. 
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #alpha creates grey scale of the geom_points
```

```{r Change point shape by class.}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

You can also set these properties manually, such as by offering a number or a colour. Let’s make all of our points blue. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

#2.19 Troubleshooting

pluss in wrong place

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
#it was at the beginning of the second line
```

2.20 Facet and panel plots

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  facet_wrap(~class,nrow=2)+
  labs(x="Engine size",y="Fuiel efficiency")

#Engine size = displ
#Fuiel efficiency = hwy
```

```{r}
ggplot(data=mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  facet_wrap(drv~cyl)+
  labs(x="Engine size",y="Fuiel efficiency")
```

Use a . if you do not want to facet in the rows or column dimension. Note what happens in this case:

```{r}
ggplot(data=mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  facet_wrap(.~cyl)+
  labs(x="Engine size",y="Fuiel efficiency")
# it orgaices by amount of cylynders and not by both cyl and drv
```
#2.21 Fitting simple lines

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))
```

```{r}
ggplot(data = mpg)+
#  geom_point(mapping = aes(x = displ, y = hwy)) # points horrible 
  geom_smooth(mapping = aes(x=displ,y=hwy))# try smooth line
```

```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x=displ,y=hwy, linetype = drv))
```

Set the group aesthetic to a categorical variable to draw multiple objects.

```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x=displ,y=hwy,group = drv))
```

Change the color of each line based on drv value
```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x=displ,y=hwy,colour= drv),
              show.legend = FALSE)
```

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  geom_smooth(mapping = aes(x=displ,y=hwy))
```

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy))+
  geom_point()+
  geom_smooth()
```

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy))+
  geom_point(mapping = aes(colour = class))+
  geom_smooth()
```

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy))+
  geom_point(mapping = aes(colour = class))+
  geom_smooth(data = filter(mpg,class=="subcompact"),se=FALSE)
```

Exercise:
1.	What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2.	Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions. Will these two graphs look different? Why/why not?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
they look the same, there is the same info, one is just a less character way of plotting


#2.22 Transformations and Stats
##2.22.1 Plotting statistics


Our first bar chart shows that more diamonds are available with high quality cuts than low quality cuts.
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut))
```

```{r}
ggplot(data = diamonds)+
  stat_count(mapping = aes(x=cut))
```
##2.22.2 Overriding defaults in ggplot2
```{r}
demo <- tribble(
  ~cut,            ~freq,
  "Fair",           1610,
  "Good",           4906,
  "Verry Good",     12082,
  "Premium",        13791,
  "Ideal",          21551
)
demo
#what is the problem with this?
```

```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")

```

```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,y=stat(prop),group = 1))
#> Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
```
#2.22.3 Plotting statistical details
```{r}
ggplot(data = diamonds)+
  stat_summary(
    mapping = aes(x=cut,y=depth),
    fun.min = min,
    fun.max = max,
    fun=median
  )
```
#2.23 Aesthetic adjustments

```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,colour = cut))
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill=cut))
```

```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill = clarity))
```


```{r}
#to alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x=cut,fill = clarity))+
  geom_bar(alpha=1/5,position = "identity")

#to color the bar outlines with no fill colour
ggplot(data = diamonds,mapping = aes(x=cut,colour = clarity))+
  geom_bar(fill=NA,position = "identity")
```

●	position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill = clarity),position = "fill")
```
●	position = "dodge" places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill = clarity),position = "dodge")
```
●	position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy),position = "jitter")
```


#Workshop 4

##Step 1: prepare your workspace
```{r}
#load tidyverse and palmerpenguins packages
library(tidyverse)
library(palmerpenguins)
```


##Step 2: Prompt for a Multi-Variable Scatter Plot
```{r}
#create a ggplot scatter plot using the penguins dataset 
#plot bill_length_mm on the x-axis and bill_depth_mm on the y-axis
#color points by species and geom_point with size 3
# add a minimal theme and clean lables
ggplot(data = penguins, mapping = aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Bill Length vs Bill Depth in Penguins",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")
```

##Step 3: Open-end comarative plotting
###Option A (Distribution Check): 
Instruct Copilot to generate a boxplot comparing flipper_length_mm across different species, faceted by island, using custom color palettes.
```{r}
#create a ggplot boxplot comparing flipper_length_mm across species from the pneguins dataset
#Facet the plot by island 
#Usea use only pastel colors
#Drop the names on the x axis and add a minimal theme

ggplot(data = penguins, mapping = aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ island) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(title = "Flipper Length Distribution by Species and Island",
       x = "Species",
       y = "Flipper Length (mm)",
       fill = "Species")
```


# step 4: verify and Document
Did it work? yes it worked, compilot had a few creative ideas of its own, i did leave some of my props open for interretation, i could have been more precis, but i think it came out good. 



