---
title: "Workshop 2"
output: html_notebook
---

#2.16	Load packages and data
```{r library}
library(dplyr)
library(ggplot2) 
library(tidyverse)
```

##using the built-in dataset 
```{r}
mpg
```

#2.17 Create your first ggplot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Engine size = displ
Fuiel efficiency = hwy

the plot shows a negative relationship between engine size and fuil efficiency

#2.18 Understand the ‘grammar of graphics’
##2.18.1 Graphing template

##2.18.2 Aesthetic mappings
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))#using size,sizes up the geom_points
#> Warning: Using size for a discrete variable is not advised. 
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #alpha creates grey scale of the geom_points
```

```{r Change point shape by class.}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

You can also set these properties manually, such as by offering a number or a colour. Let’s make all of our points blue. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

#2.19 Troubleshooting

pluss in wrong place

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
#it was at the beginning of the second line
```

2.20 Facet and panel plots

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  facet_wrap(~class,nrow=2)+
  labs(x="Engine size",y="Fuiel efficiency")

#Engine size = displ
#Fuiel efficiency = hwy
```

```{r}
ggplot(data=mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  facet_wrap(drv~cyl)+
  labs(x="Engine size",y="Fuiel efficiency")
```

Use a . if you do not want to facet in the rows or column dimension. Note what happens in this case:

```{r}
ggplot(data=mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  facet_wrap(.~cyl)+
  labs(x="Engine size",y="Fuiel efficiency")
# it orgaices by amount of cylynders and not by both cyl and drv
```
#2.21 Fitting simple lines

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))
```

```{r}
ggplot(data = mpg)+
#  geom_point(mapping = aes(x = displ, y = hwy)) # points horrible 
  geom_smooth(mapping = aes(x=displ,y=hwy))# try smooth line
```

```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x=displ,y=hwy, linetype = drv))
```

Set the group aesthetic to a categorical variable to draw multiple objects.

```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x=displ,y=hwy,group = drv))
```

Change the color of each line based on drv value
```{r}
ggplot(data = mpg)+
  geom_smooth(mapping = aes(x=displ,y=hwy,colour= drv),
              show.legend = FALSE)
```

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy))+
  geom_smooth(mapping = aes(x=displ,y=hwy))
```

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy))+
  geom_point()+
  geom_smooth()
```

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy))+
  geom_point(mapping = aes(colour = class))+
  geom_smooth()
```

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy))+
  geom_point(mapping = aes(colour = class))+
  geom_smooth(data = filter(mpg,class=="subcompact"),se=FALSE)
```

Exercise:
1.	What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2.	Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions. Will these two graphs look different? Why/why not?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
they look the same, there is the same info, one is just a less character way of plotting


#2.22 Transformations and Stats
##2.22.1 Plotting statistics


Our first bar chart shows that more diamonds are available with high quality cuts than low quality cuts.
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut))
```

```{r}
ggplot(data = diamonds)+
  stat_count(mapping = aes(x=cut))
```
##2.22.2 Overriding defaults in ggplot2
```{r}
demo <- tribble(
  ~cut,            ~freq,
  "Fair",           1610,
  "Good",           4906,
  "Verry Good",     12082,
  "Premium",        13791,
  "Ideal",          21551
)
demo
#what is the problem with this?
```

```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")

```

```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,y=stat(prop),group = 1))
#> Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
```
#2.22.3 Plotting statistical details
```{r}
ggplot(data = diamonds)+
  stat_summary(
    mapping = aes(x=cut,y=depth),
    fun.min = min,
    fun.max = max,
    fun=median
  )
```
#2.23 Aesthetic adjustments

```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,colour = cut))
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill=cut))
```

```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill = clarity))
```


```{r}
#to alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x=cut,fill = clarity))+
  geom_bar(alpha=1/5,position = "identity")

#to color the bar outlines with no fill colour
ggplot(data = diamonds,mapping = aes(x=cut,colour = clarity))+
  geom_bar(fill=NA,position = "identity")
```

●	position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill = clarity),position = "fill")
```
●	position = "dodge" places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x=cut,fill = clarity),position = "dodge")
```
●	position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy),position = "jitter")
```


